---
title: "Levier"
format:
  dashboard:
    orientation: columns
    nav-buttons: []
---

```{r}
#| context: setup
#| include: false

library(leaflet)
library(htmlwidgets)

library(dplyr)
library(jsonlite)
library(ggplot2)
library(plotly)
library(DT)
library(forcats)

levier_geojson <- paste(readLines("data/levier.geojson", warn = FALSE), collapse = "\n")
parcelles_geojson <- paste(readLines("data/parcelles.geojson", warn = FALSE), collapse = "\n")
gbif_geojson <- paste(readLines("data/gbif_occurrences_levier.geojson", warn = FALSE), collapse = "\n")

# species -> common name lookup (built by code/fetch_vernacular_names.R), used
# to label the sidebar dropdowns with common names instead of Latin names
vernacular <- read.csv("data/gbif_vernacular_names.csv", stringsAsFactors = FALSE) %>%
  filter(!is.na(commonName), commonName != "")
vernacular_json <- toJSON(setNames(as.list(vernacular$commonName), vernacular$species), auto_unbox = TRUE)

# GeoJSON is loaded client-side (via L.geoJSON) rather than through sf, so that
# each polygon keeps its `feature.properties` and can drive the click handler below.
map_js <- r"(
function(el, x) {
  var map = this;

  var osm = L.tileLayer('https://{s}.tile.openstreetmap.org/{z}/{x}/{y}.png', {
    maxZoom: 19,
    attribution: '&copy; OpenStreetMap contributors'
  }).addTo(map);

  var esri = L.tileLayer('https://server.arcgisonline.com/ArcGIS/rest/services/World_Imagery/MapServer/tile/{z}/{y}/{x}', {
    maxZoom: 19,
    attribution: 'Tiles &copy; Esri, Maxar, Earthstar Geographics'
  });

  var ignOrtho = L.tileLayer('https://data.geopf.fr/wmts?SERVICE=WMTS&VERSION=1.0.0&REQUEST=GetTile&LAYER=ORTHOIMAGERY.ORTHOPHOTOS&STYLE=normal&TILEMATRIXSET=PM&TILEMATRIX={z}&TILEROW={y}&TILECOL={x}&FORMAT=image/jpeg', {
    maxZoom: 19,
    minZoom: 0,
    attribution: '&copy; IGN-F/Geoportail'
  });

  var ignCadastre = L.tileLayer('https://data.geopf.fr/wmts?SERVICE=WMTS&VERSION=1.0.0&REQUEST=GetTile&LAYER=CADASTRALPARCELS.PARCELS&STYLE=bdparcellaire&TILEMATRIXSET=PM&TILEMATRIX={z}&TILEROW={y}&TILECOL={x}&FORMAT=image/png', {
    maxZoom: 19,
    minZoom: 0,
    attribution: '&copy; IGN-F/Geoportail, DGFiP'
  });

  var communeLayer = L.geoJSON(__LEVIER_GEOJSON__, {
    style: { fillColor: 'transparent', color: '#3182bd', weight: 2, fillOpacity: 0 }
  }).addTo(map);

  var parcellesLayer = L.geoJSON(__PARCELLES_GEOJSON__, {
    style: { fillColor: '#fdae6b', fillOpacity: 0.3, color: '#e6550d', weight: 1 },
    onEachFeature: function(feature, layer) {
      layer.on('click', function(e) {
        var p = feature.properties;
        var rows = '';
        Object.keys(p).forEach(function(k) {
          rows += '<tr><th>' + k + '</th><td>' + p[k] + '</td></tr>';
        });
        document.getElementById('parcelle-info').innerHTML =
          '<table class="table table-sm table-striped">' + rows + '</table>';
      });
      layer.on('mouseover', function(e) { e.target.setStyle({ weight: 3, color: '#000000' }); });
      layer.on('mouseout', function(e) { parcellesLayer.resetStyle(e.target); });
    }
  }).addTo(map);

  map.fitBounds(communeLayer.getBounds());

  var layersControl = L.control.layers(
    { 'OpenStreetMap': osm, 'Imagerie (ESRI)': esri, 'Photos aeriennes (IGN)': ignOrtho },
    { 'Commune': communeLayer, 'Parcelles': parcellesLayer, 'Parcelles cadastrales (IGN)': ignCadastre },
    { collapsed: false }
  ).addTo(map);

  // -- GBIF occurrences: embedded directly (like the other layers) so the
  // dashboard works when opened as a plain file, without a web server --
  var gbifLayer = L.layerGroup();
  var gbifMarkers = [];

  var vernacular = __VERNACULAR_JSON__;
  var kingdomLabels = { Animalia: 'Animaux', Plantae: 'Plantes', Fungi: 'Champignons', Protozoa: 'Protozoaires' };

  function commonName(species) { return species ? vernacular[species] : undefined; }

  function speciesLabel(species) {
    var common = commonName(species);
    if (!common) return species;
    return common.charAt(0).toUpperCase() + common.slice(1) + ' (' + species + ')';
  }

  function gbifPopup(p) {
    return '<strong>' + speciesLabel(p.species || p.scientificName) + '</strong><br>' +
      (p.eventDate ? p.eventDate + '<br>' : '') +
      (p.recordedBy ? 'Observateur : ' + p.recordedBy + '<br>' : '') +
      (p.datasetTitle ? '<span class="text-muted">' + p.datasetTitle + '</span>' : '');
  }

  function gbifAttributeTable(p) {
    var rows = '';
    if (p.species) rows += '<tr><th>Nom commun</th><td>' + (commonName(p.species) || '&mdash;') + '</td></tr>';
    Object.keys(p).forEach(function(k) {
      var val = p[k];
      if (val === null || val === undefined || val === '') return;
      rows += '<tr><th>' + k + '</th><td>' + val + '</td></tr>';
    });
    return '<table class="table table-sm table-striped">' + rows + '</table>';
  }

  function updateGbifCount(shown) {
    var el = document.getElementById('gbif-count');
    if (el) el.textContent = shown + ' / ' + gbifMarkers.length + ' observations affichees';
  }

  function applyGbifFilter() {
    var kingdom = document.getElementById('gbif-kingdom').value;
    var species = document.getElementById('gbif-species').value;
    var shown = 0;
    gbifMarkers.forEach(function(m) {
      var p = m.feature.properties;
      var match = (!kingdom || p.kingdom === kingdom) && (!species || p.species === species);
      if (match) {
        if (!gbifLayer.hasLayer(m)) gbifLayer.addLayer(m);
        shown++;
      } else if (gbifLayer.hasLayer(m)) {
        gbifLayer.removeLayer(m);
      }
    });
    updateGbifCount(shown);
  }

  function populateSelect(id, values, placeholder, labelFn) {
    var select = document.getElementById(id);
    select.innerHTML = '';
    var opt0 = document.createElement('option');
    opt0.value = '';
    opt0.textContent = placeholder;
    select.appendChild(opt0);
    values.forEach(function(v) {
      var opt = document.createElement('option');
      opt.value = v;
      opt.textContent = labelFn ? labelFn(v) : v;
      select.appendChild(opt);
    });
  }

  function refreshSpeciesOptions() {
    var kingdom = document.getElementById('gbif-kingdom').value;
    var species = gbifMarkers
      .map(function(m) { return m.feature.properties; })
      .filter(function(p) { return !kingdom || p.kingdom === kingdom; })
      .map(function(p) { return p.species; })
      .filter(function(s) { return !!s; });
    var uniqueSpecies = Array.from(new Set(species)).sort(function(a, b) {
      return speciesLabel(a).localeCompare(speciesLabel(b), 'fr');
    });
    populateSelect('gbif-species', uniqueSpecies, 'Toutes les especes', speciesLabel);
  }

  (function(data) {
    data.features.forEach(function(f) {
      if (!f.geometry || !f.geometry.coordinates) return;
      var latlng = [f.geometry.coordinates[1], f.geometry.coordinates[0]];
      var marker = L.circleMarker(latlng, {
        radius: 4, weight: 1, color: '#238b45', fillColor: '#74c476', fillOpacity: 0.7
      });
      marker.feature = f;
      marker.bindPopup(gbifPopup(f.properties));
      marker.on('click', function() {
        var info = document.getElementById('gbif-info');
        if (info) info.innerHTML = gbifAttributeTable(f.properties);
      });
      gbifMarkers.push(marker);
      gbifLayer.addLayer(marker);
    });

    gbifLayer.addTo(map);
    layersControl.addOverlay(gbifLayer, 'Observations GBIF');

    var kingdoms = Array.from(new Set(
      gbifMarkers.map(function(m) { return m.feature.properties.kingdom; }).filter(function(k) { return !!k; })
    )).sort(function(a, b) {
      return (kingdomLabels[a] || a).localeCompare(kingdomLabels[b] || b, 'fr');
    });
    populateSelect('gbif-kingdom', kingdoms, 'Tous les regnes', function(k) { return kingdomLabels[k] || k; });
    refreshSpeciesOptions();
    updateGbifCount(gbifMarkers.length);

    document.getElementById('gbif-kingdom').addEventListener('change', function() {
      refreshSpeciesOptions();
      applyGbifFilter();
    });
    document.getElementById('gbif-species').addEventListener('change', applyGbifFilter);
    document.getElementById('gbif-reset').addEventListener('click', function() {
      document.getElementById('gbif-kingdom').value = '';
      refreshSpeciesOptions();
      document.getElementById('gbif-species').value = '';
      applyGbifFilter();
    });
  })(__GBIF_GEOJSON__);
}
)"

map_js <- gsub("__LEVIER_GEOJSON__", levier_geojson, map_js, fixed = TRUE)
map_js <- gsub("__PARCELLES_GEOJSON__", parcelles_geojson, map_js, fixed = TRUE)
map_js <- gsub("__GBIF_GEOJSON__", gbif_geojson, map_js, fixed = TRUE)
map_js <- gsub("__VERNACULAR_JSON__", vernacular_json, map_js, fixed = TRUE)

# Data frame form of the GBIF occurrences, for the summary tables/charts on
# the second page (as opposed to the raw GeoJSON text used by the map above).
gbif_df <- read.csv("data/gbif_occurrences_levier.csv", stringsAsFactors = FALSE) |>
  left_join(vernacular %>% select(species, commonName), by = "species") |>
  mutate(
    kingdom_fr = recode(kingdom,
      Animalia = "Animaux", Plantae = "Plantes", Fungi = "Champignons", Protozoa = "Protozoaires",
      .missing = "Inconnu"
    ),
    common_name_cap = ifelse(
      !is.na(commonName) & commonName != "",
      paste0(toupper(substr(commonName, 1, 1)), substr(commonName, 2, nchar(commonName))),
      NA_character_
    ),
    display_name = ifelse(!is.na(common_name_cap), paste0(common_name_cap, " (", species, ")"), species)
  )
```

# Carte

## {.sidebar}

**Observations GBIF**

<div class="mb-2">
<label for="gbif-kingdom" class="form-label">Regne</label>
<select id="gbif-kingdom" class="form-select form-select-sm">
<option value="">Chargement...</option>
</select>
</div>

<div class="mb-2">
<label for="gbif-species" class="form-label">Espece</label>
<select id="gbif-species" class="form-select form-select-sm">
<option value="">Chargement...</option>
</select>
</div>

<button id="gbif-reset" class="btn btn-sm btn-outline-secondary">Reinitialiser</button>

<p class="small text-muted mt-2" id="gbif-count">Chargement des observations...</p>

## Column {width=65%}

```{r}
#| title: "Carte"

leaflet(options = leafletOptions(preferCanvas = TRUE)) |>
  onRender(map_js)
```

## Column {width=35%}

### Row {height=50%}

::: {.card title="Informations parcelle"}

<div id="parcelle-info">
<p><em>Cliquez sur une parcelle pour afficher ses informations.</em></p>
</div>

:::

### Row {height=50%}

::: {.card title="Observation GBIF"}

<div id="gbif-info">
<p><em>Cliquez sur un point d'observation pour afficher ses informations.</em></p>
</div>

:::

# Resume GBIF {orientation="rows"}

## Row {height=15%}

```{r}
#| content: valuebox
#| title: "Observations"
list(value = format(nrow(gbif_df), big.mark = " "), icon = "binoculars", color = "primary")
```

```{r}
#| content: valuebox
#| title: "Especes"
list(value = n_distinct(gbif_df$species[gbif_df$species != ""]), icon = "collection", color = "success")
```

```{r}
#| content: valuebox
#| title: "Familles"
list(value = n_distinct(gbif_df$family[gbif_df$family != ""]), icon = "diagram-3-fill", color = "info")
```

```{r}
#| content: valuebox
#| title: "Periode"
list(
  value = paste0(min(gbif_df$year, na.rm = TRUE), " - ", max(gbif_df$year, na.rm = TRUE)),
  icon = "calendar3",
  color = "warning"
)
```

## Row {height=40%}

### Column {width=50%}

```{r}
#| title: "Especes les plus observees"

top_species <- gbif_df |>
  filter(!is.na(species), species != "") |>
  count(display_name, name = "n", sort = TRUE) |>
  slice_head(n = 15)

p <- ggplot(top_species, aes(x = n, y = fct_reorder(display_name, n))) +
  geom_col(fill = "#238b45") +
  labs(x = "Observations", y = NULL) +
  theme_minimal(base_size = 11)

ggplotly(p, tooltip = c("x", "y")) |>
  layout(margin = list(l = 220))
```

### Column {width=50%}

```{r}
#| title: "Repartition par regne"

kingdom_counts <- gbif_df |>
  filter(!is.na(kingdom_fr)) |>
  count(kingdom_fr, name = "n", sort = TRUE)

p <- ggplot(kingdom_counts, aes(x = fct_reorder(kingdom_fr, n), y = n, fill = kingdom_fr)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(x = NULL, y = "Observations") +
  theme_minimal(base_size = 11)

ggplotly(p, tooltip = c("x", "y"))
```

## Row {height=20%}

```{r}
#| title: "Observations par annee"

year_counts <- gbif_df |>
  filter(!is.na(year)) |>
  count(year, name = "n")

p <- ggplot(year_counts, aes(x = year, y = n)) +
  geom_col(fill = "#3182bd") +
  labs(x = NULL, y = "Observations") +
  theme_minimal(base_size = 11)

ggplotly(p, tooltip = c("x", "y"))
```

## Row {height=25%}

```{r}
#| title: "Toutes les especes observees"

species_table <- gbif_df |>
  filter(!is.na(species), species != "") |>
  group_by(species, common_name_cap, kingdom_fr, family) |>
  summarise(observations = n(), .groups = "drop") |>
  arrange(desc(observations)) |>
  transmute(
    `Nom commun` = ifelse(is.na(common_name_cap), "—", common_name_cap),
    `Nom scientifique` = species,
    `Regne` = kingdom_fr,
    `Famille` = ifelse(is.na(family) | family == "", "—", family),
    `Observations` = observations
  )

datatable(
  species_table,
  rownames = FALSE,
  options = list(pageLength = 10, order = list(list(4, "desc")))
)
```
